---
title: "waiind: A High-Frequency GDP Indicator for Switzerland"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{waiind: A High-Frequency GDP Indicator for Switzerland}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(waiind)
```

## What is the WAI?

The Weekly Activity Index (WAI) is a high-frequency GDP indicator for
Switzerland: a single dynamic factor, extracted from a Bayesian
mixed-frequency dynamic factor model, that is directly interpretable as
the week-on-week growth rate of Swiss real GDP. It is built to combine
two kinds of data:

- **Conventional macroeconomic indicators** — surveys, prices, trade,
  purchasing manager indices — published monthly or quarterly, and
- **Alternative high-frequency data** — payment and transaction volumes,
  mobility indicators, search-trend indices, traffic and flight counts —
  available daily or weekly, often with very short histories and
  irregular publication lags.

Unlike unsupervised approaches that extract a factor from high-frequency
data alone (e.g. principal components), the WAI is *supervised*: an
identification restriction anchors the factor to observed GDP growth, so
it is a direct high-frequency proxy for GDP rather than a generic
activity index that merely correlates with it.

The methodology, full empirical results, and a real-time forecast
evaluation are described in Kronenberg (2026); see `?waiind` for the
full citation and the underlying multi-factor framework of Eckert,
Kronenberg, Mikosch & Neuwirth (2025). This vignette covers the
package's user-facing workflow only.

## The model, briefly

`hfdfm()` estimates a single latent weekly factor $f_t$ whose
(distributed-lag, temporally aggregated) measurement equation links it
to every input series, and whose autoregressive state equation includes
**stochastic volatility** so the variance of common shocks can rise
during crises instead of being smoothed away. Three features make this
work with real-world mixed-frequency data:

- **Data augmentation** treats missing and lower-frequency observations
  as latent states, so a series that starts in 2020 and updates weekly,
  another that starts in 1990 and updates monthly, and quarterly GDP
  itself can all enter the same state-space model.
- **Stochastic volatility** in the factor state equation lets abrupt
  swings (e.g. the COVID-19 pandemic) show up as genuinely higher
  volatility rather than being averaged away.
- **Quasi-differencing** of the measurement equation removes serial
  correlation in each series' idiosyncratic errors, so persistent
  indicator-specific noise doesn't get misattributed to the common
  factor.

Identification fixes the factor's loading on the target series (GDP) to
one and shrinks its measurement error toward zero, so the factor tracks
GDP growth closely by construction — see `?hfdfm` for the details and
references, and `vignette("waiind")`'s "Real-time GDP vintages" section
below for how the target series itself is constructed for live use.

## A minimal worked example

The package ships a curated, model-ready dataset,
`data_ch_dataset_test`, so a small demonstration model can be estimated
without any external data. Real analyses use the full indicator set and
a much longer chain (the accompanying paper uses `length_sample = 5000`
after a burn-in of 1000); here we use a tiny subset and a short chain
purely so this vignette builds quickly.

```{r fit-model}
data(data_ch_dataset_test)

target <- "ch.seco.gdp.real.gdp.ssa"
flows <- lapply(
  data_ch_dataset_test$flows[c(target, "SWISSMI", "traffic_PW")],
  stats::window, start = 2018
)
stocks <- lapply(data_ch_dataset_test$stocks[1:2], stats::window, start = 2018)

set.seed(1)
fit <- hfdfm(
  flows = flows, stocks = stocks, target = target,
  length_sample = 200, burn_in = 50
)

class(fit)
names(fit)
```

`fit$factor` is the annualized weekly growth rate implied by the model —
the WAI itself — and `fit$nowcast` is that same information aggregated
back up to the target's own (here quarterly) frequency:

```{r plot-factor, fig.width = 6, fig.height = 4}
plot(fit$factor, ylab = "WAI (annualized %, weekly)", xlab = NULL)
```

```{r nowcast}
fit$nowcast
```

`fit$index` gives the cumulated activity level (rebased to 100 at the
start of the estimation window), and `fit$pars` holds the posterior
means of the model parameters, including the factor loadings
(`fit$pars$lambda`) — the target series' loading is fixed at 1 by
construction:

```{r loadings}
fit$pars$lambda[which(fit$inventory$key == target)]
```

## Real-time GDP vintages

The full `data_ch_dataset` deliberately ships *without* a GDP target
series: real analyses inject GDP at runtime from the real-time vintage
database, which ships with the package and needs no configuration:

```{r vintages}
vintages <- get_real_time_gdp_vintages("quarterly")
dim(vintages)
```

`cut_data_real_time()` truncates a dataset to what would have been
observable at a given date, substituting the appropriate GDP vintage for
the target series — the building block for the paper's real-time
out-of-sample evaluation:

```{r cut-real-time}
dat_rt <- cut_data_real_time(data_ch_dataset_test, current_date = 2024.5,
                             GDP_gr_vintages = vintages)
tail(time(dat_rt$flows[[target]]))
```

## Beyond this vignette

The functions used in the accompanying analysis pipeline are exported
but need either the full (non-shipped) indicator set or previously
saved model fits, so they aren't run here:

```{r pipeline, eval = FALSE}
# Estimate the full WAI and optionally save the fit:
fit_full <- run_wai_adj(
  flows = dat$flows, stocks = dat$stocks, target = target,
  date = 2024.5, dataset_used = "full_RT", output_dir = "fits/updated"
)

# AR(1) benchmark, and the Diebold-Mariano test used to compare them:
fit_ar <- run_ar(flows = dat$flows, stocks = dat$stocks, target = target,
                 date = 2024.5, dataset_used = "full_RT")
dm_test_modified(errors_wai, errors_ar)
```

For the in-sample and out-of-sample evaluation tables and plots shown in
the paper (correlation heatmaps, relative RMSE/MAE tables against the
SECO-WEA, F-CURVE, SECO-SEC, SNB-BCI and KOF-BARO benchmarks), see the
`analysis/5_plots/` scripts in the package's source repository, which
call `get_combined_cor_table()`, `get_insample_fit_table()`,
`create_rel_error_tables()` and related functions documented under
`?get_combined_cor_table`.
